GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-11-17 figs/GSE158422_rna_int.Rmd
force use of EPIC annotation for GSAmeth

@@

suppressPackageStartupMessages({
  library("DESeq2")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("HGNChelper")
  library("gplots")
})

@@

## Load annotations

@@

load("GSE158422_gmea.Rdata")

@@

Pathway enrichment with mitch.

@@

y <- mitch_import(dge, DEtype="deseq2")

res <- mitch_calc(y, genesets=gs_symbols, priority="effect",cores=CORES)

@@

```{r,overlap1}

load("GSE158422_gmea.Rdata")

head(mres)

head(alres)

head(lares)

msig_up <- subset(msig,s.dist>0)$set
msig_dn <- subset(msig,s.dist<0)$set

lasig_up <- rownames(subset(lasig,t_median>0))
lasig_dn <- rownames(subset(lasig,t_median<0))

alsig_up <- rownames(subset(alsig,t_median>0))
alsig_dn <- rownames(subset(alsig,t_median<0))

v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

@@

## Load promoter methylation enrichment data and calculate overlap

```{r,overlap2}

head(alpres)

head(lapres)

lapsig_up <- rownames(subset(lapsig,t_median>0))
lapsig_dn <- rownames(subset(lapsig,t_median<0))

alpsig_up <- rownames(subset(alpsig,t_median>0))
alpsig_dn <- rownames(subset(alpsig,t_median<0))

v1 <- list("LAP up"=lapsig_up, "LAP dn"=lapsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("ALP up"=alpsig_up,"ALP dn"=alpsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

>>>>

Let's take a look at how pathway methylation associates with gene expression.

Partition methylation analysis between promoter and body.

Compare to GSAmeth.

@@

suppressPackageStartupMessages({
  library("DESeq2")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("HGNChelper")
  library("gplots")
})

@@

## Load annotations and methylation data

@@

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))
summary(unlist(lapply(sets,length)))

design <- readRDS("GSE158422_design.rds")
mval <- readRDS("GSE158422_mx.rds")
dm <- readRDS("GSE158422_dm.rds")

@@

RNA expression pathway enrichment with mitch.

@@

rna <- mitch_import(dge, DEtype="deseq2")

res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=32)

@@

```{r,lamfunc}

# LAM aggregate
agg <- function(dm,cores=1) {
  dm <- dm[which(dm$UCSC_RefGene_Name!=""),]
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    top <- tstats[order(-abs(tstats))][1]
    if ( length(tstats) > 2 ) {
      ttest <- t.test(tstats)
      pval <- ttest$p.value
    } else {
      pval = 1
    }
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,
      "median"=mymedian, "top"=top, "pval"=pval)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,"pval"])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$fdr <- p.adjust(gme_res_df$pval)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

@@

Here we will look at parallel analysis of gene body and promoter.

```{r,promoterbody}

table(dm$Regulatory_Feature_Group)

# all gene
dmagg <- agg(dm=dm,cores=16)
str(dmagg)
head(dmagg$gme_res_df)
dim(dm)
dim(dmagg$gme_res_df)
mm <- dmagg$gme_res_df[,"mean",drop=FALSE]
mmres <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=32)
mmres_up <- subset(mmres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmres_dn <- subset(mmres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmres_up)
nrow(mmres_dn)
head(mmres_up,20)
head(mmres_dn,20)

# promoter
dmpr <- dm[grep("Promoter",dm$Regulatory_Feature_Group),]
dmpragg <- agg(dm=dmpr,cores=16)
str(dmpragg)
head(dmpragg$gme_res_df)
dim(dmpr)
dim(dmpragg$gme_res_df)
mmpr <- dmpragg$gme_res_df[,"mean",drop=FALSE]
mmprres <- mitch_calc(mmpr, genesets=gs_symbols, priority="effect",cores=32)
mmprres_up <- subset(mmprres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmprres_dn <- subset(mmprres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmprres_up)
nrow(mmprres_dn)
head(mmprres_up,20)
head(mmprres_dn,20)

# ex-promoter
dmx <- dm[grep("Promoter",dm$Regulatory_Feature_Group,invert=TRUE),]
dmxagg <- agg(dm=dmx,cores=16)
str(dmxagg)
head(dmxagg$gme_res_df)
dim(dmx)
dim(dmxagg$gme_res_df)
mmx <- dmxagg$gme_res_df[,"mean",drop=FALSE]
mmxres <- mitch_calc(mmx, genesets=gs_symbols, priority="effect",cores=32)
mmxres_up <- subset(mmxres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmxres_dn <- subset(mmxres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmxres_up)
nrow(mmxres_dn)
head(mmxres_up,20)
head(mmxres_dn,20)

# combined methylation
mmm <- merge(mmpr,mmx,by=0)
rownames(mmm) <- mmm$Row.names
mmm$Row.names = NULL
colnames(mmm) <- c("pr","x")
mmmres <- mitch_calc(mmm, genesets=gs_symbols, priority="effect",cores=32)
mmmres_sig <- subset(mmmres$enrichment_result, p.adjustMANOVA < 0.05)
head(mmmres_sig,20)
table(sign(mmmres_sig$s.pr) == sign(mmmres_sig$s.x))

## RNA expression and methylation enrichment

```{r,rnaoverlap2}

# three way
mmrna <- merge(mmm,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("pr","x","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.x))
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
mtop <- head(mmrnares_sig,30)
mtop <- mtop[,c("set","s.pr","s.x","s.rna")]
rownames(mtop) <- gsub("REACTOME_","",mtop$set)
rownames(mtop) <- gsub("_"," ",rownames(mtop))
mtop$set=NULL
colnames(mtop) <- c("promoter","non-promoter","RNA")
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)

# rna vs promoter
mmrna <- merge(mmpr,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("pr","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.pr)
summary(mylm)
mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.pr)
summary(mylm2)
plot(mmrnares$enrichment_result$s.pr,mmrnares$enrichment_result$s.rna,xlab="promoter",ylab="RNA",main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mmrnares_sig$s.pr,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

# rna vs expromoter
mmrna <- merge(mmx,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("x","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.x)
summary(mylm)
mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.x)
summary(mylm2)
plot(mmrnares$enrichment_result$s.x,mmrnares$enrichment_result$s.rna,xlab="ex-promoter",ylab="RNA",main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mmrnares_sig$s.x,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

Now look at GSAmeth.

```{r,rnagsa}

# all meth
sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn)
nrow(gsig)

# all meth
sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn) 
nrow(gsig)

### 2023-11-17 figs/GSE158422_sensitivity.Rmd
force use of EPIC annotation for GSAmeth

dim(dm)
nrow(subset(dm,adj.P.Val<0.05 & logFC>0))
nrow(subset(dm,adj.P.Val<0.05 & logFC<0))

@@

sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
nsigs <- lapply(sizes, function(z) {
  dmsig <- head(dm,z)
  dmsigup <- rownames(subset(dmsig,logFC>0))
  dmsigdn <- rownames(subset(dmsig,logFC<0))
  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  nsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(nsig)
})

@@

#
FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
ngsig <- lapply(FCLIM,function(F) {
  sigup <- rownames(subset(dm,logFC > F & adj.P.Val < 0.05))
  sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  ngsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(ngsig)
})

@@

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
toc() # 12.9, 12.6, 11.5

@@

gsres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    #mval downsample
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval2,design2,myann)
    if ( is.na(dm2$logFC[1]) ) {
      RES=c("TOT"=0,"TP"=0,"FP"=0,"FN"=length(rownames(gsig)) )
    } else {
      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
      rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
      rownames(gsadn2) <- paste(rownames(gsadn2),"DN")
      gsig_up2 <- subset(gsaup2,FDR<0.05)
      gsig_dn2 <- subset(gsadn2,FDR<0.05)
      gsig2 <- rbind(gsig_up2,gsig_dn2)
      TOT=nrow(gsig2)
      TP=length(which(rownames(gsig2) %in% rownames(gsig)))
      FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
      FN=length(which(!rownames(gsig) %in% rownames(gsig2)))
      RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    }
    RES
  }, mc.cores=5)
  do.call(rbind,dsres)
})

>>>>

dim(dm)
nrow(subset(dm,adj.P.Val<0.05 & logFC>0))
nrow(subset(dm,adj.P.Val<0.05 & logFC<0))
saveRDS(dm,"GSE158422_dm.rds")

@@

sizes=c(1000,5000,10000,20000,50000,100000,200000,300000,350000,400000,450000,500000)
nsigs <- lapply(sizes, function(z) {
  dmsig <- head(dm,z)
  dmsigup <- rownames(subset(dmsig,logFC>0))
  dmsigdn <- rownames(subset(dmsig,logFC<0))
  gsaup <- gsameth(sig.cpg=dmsigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=dmsigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  nsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(nsig)
})

@@

#
FCLIM <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
ngsig <- lapply(FCLIM,function(F) {
  sigup <- rownames(subset(dm,logFC > F & adj.P.Val < 0.05))
  sigdn <- rownames(subset(dm,logFC < -F & adj.P.Val < 0.05))
  gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
  rownames(gsaup) <- paste(rownames(gsaup),"UP")
  gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
  rownames(gsadn) <- paste(rownames(gsadn),"DN")
  ngsig <- rownames(subset(rbind(gsaup,gsadn),FDR<0.05))
  return(ngsig)
})

@@

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
toc() # 12.9, 12.6, 11.5

@@

gsres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    #mval downsample
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval2,design2,myann)
    if ( is.na(dm2$logFC[1]) ) {
      RES=c("TOT"=0,"TP"=0,"FP"=0,"FN"=length(rownames(gsig)) )
    } else {
      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez, array.type="EPIC")
      rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez, array.type="EPIC")
      rownames(gsadn2) <- paste(rownames(gsadn2),"DN")
      gsig_up2 <- subset(gsaup2,FDR<0.05)
      gsig_dn2 <- subset(gsadn2,FDR<0.05)
      gsig2 <- rbind(gsig_up2,gsig_dn2)
      TOT=nrow(gsig2)
      TP=length(which(rownames(gsig2) %in% rownames(gsig)))
      FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
      FN=length(which(!rownames(gsig) %in% rownames(gsig2)))
      RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    }
    RES
  }, mc.cores=5)
  do.call(rbind,dsres)
})
